A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-17 tex/sec2-outline.tex
[ tex ] Use blackboard font for annotation (in raw terms)
[ tex ] Fix mistakes in Anno
[ tex ] Style raw terms

\section{The Key Ideas} \label{sec:key-ideas}
\begin{figure}
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A}$ & A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A}$ &A raw term $t$ checks against a type $A$ under $\Gamma$
  \end{tabular}
  \egroup
  \centering
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \syn A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \chk B \\ A = B}{\Gamma \vdash \isTerm{t : B}\syn A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \syn B \\ A = B}{\Gamma \vdash \isTerm{t} \chk A}\,\ChkRule{Sub}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \syn A \bto B \\ \Gamma \vdash \isTerm{u} \chk A}{\Gamma \vdash \isTerm{t\;u} \syn B}\,\SynRule{App}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \chk B}{\Gamma \vdash \isTerm{\lam{x}t} \chk A \bto B}\,\ChkRule{Abs}
  \end{mathpar}
  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

>>>>

\section{The Key Ideas} \label{sec:key-ideas}
\begin{figure}
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A}$ & A raw term $\isTerm{t}$ synthesises a type $A$ under $\Gamma$ \\
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A}$ & A raw term $\isTerm{t}$ checks against a type $A$ under $\Gamma$
  \end{tabular}
  \egroup
  \centering
  \begin{mathpar}
    \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \syn A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \chk A}{\Gamma \vdash \isTerm{t \annotate A}\syn A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \syn B \\ B = A}{\Gamma \vdash \isTerm{t} \chk A}\,\ChkRule{Sub}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \syn A \bto B \\ \Gamma \vdash \isTerm{u} \chk A}{\Gamma \vdash \isTerm{t\;u} \syn B}\,\SynRule{App}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \chk B}{\Gamma \vdash \isTerm{\lam{x}t} \chk A \bto B}\,\ChkRule{Abs}
  \end{mathpar}
  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

### 2023-06-17 tex/sec3-defs.tex
[ tex ] Make some figures more compact
[ tex ] Make raw term formation rules in one line
[ tex ] Use blackboard font for annotation (in raw terms)
[ tex ] Fix mistakes in Anno
[ tex ] Style raw terms

\begin{figure}
  \begin{minipage}[b]{.6\textwidth}
    \centering
    \small
    \judgbox{\Theta|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Theta$}
    \begin{mathpar}
      \inferrule{\Theta \ni X_i}{\Theta |-_\Sigma X_i} \and
    \inferrule{\Theta |-_{\Sigma} A_1 \\ \cdots \\ \Theta |-_{\Sigma} A_{\arity(o)}}{\Theta |-_{\Sigma} \tyOp_o(A_1, \ldots, A_{\arity(o)})}
    \end{mathpar}
    \caption{Simple types}
    \label{fig:simple-type}
  \end{minipage}
  \begin{minipage}[b]{.35\textwidth}
    \centering
    \small
    \judgbox{\Theta |-_{\Sigma} \Gamma}{}
    \begin{mathpar}
      \inferrule{ }{\Theta |-_{\Sigma} \cdot }\and
      \inferrule{\Theta |-_{\Sigma} A \\ \Theta|-_{\Sigma} \Gamma}{\Theta |-_{\Sigma} \Gamma, A}
    \end{mathpar}
    \caption{Contexts}
  \label{fig:simple-context}
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \centering
  \small
  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}}{$t$ is a raw term for the signature $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}}{|-_{\Sigma, \Omega} \isTerm{t \annote A}}\,\Rule{Anno}
    \\
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}}
    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.t_1; \ldots; \vec{x}_n.t_n)}}\,\Rule{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms}
\end{figure}

\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $t$ has a type $A$ (without type variables) under $\Gamma$ for the signature $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
    \and
    \inferrule{\Gamma |- \isTerm{t} : A}{\Gamma |- (\isTerm{t \annote A}) : A}\,\Rule{Anno}
    \and
    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\,\Rule{Op}
    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Typing rules}
  \label{fig:extrinsic-typing}
\end{figure}

@@

\begin{figure}
  \centering
  \small
  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$t$ is a raw term for $(\Sigma, \erase{\Omega})$ in mode $d$ for  $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annote A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
  \end{mathpar}
  \begin{mathpar}
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\dir{d}}\,\Rule{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^{\dir{d}}$ in $\Omega$}
  \end{mathpar}
  \caption{Preprocessed raw terms, I}
\end{figure}

\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $t$ is of type $A$ (without type variables) in mode $d$ under $\Gamma$}
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annote B}) :^{\syn}  A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ A = B}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
      \cdots \\
    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} :^{\dir{d}} \simsub{A}{\rho}} \,\mathsf{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^\dir{d}$ in $\Omega$}
  \end{mathpar}
Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\chk}}$
  \caption{Bidirectional typing rules}
\end{figure}

>>>>

\begin{figure}
  \begin{minipage}[b]{.6\textwidth}
    \centering
    \small
    \judgbox{\Theta|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Theta$}
    \begin{mathpar}
      \inferrule{\Theta \ni X_i}{\Theta |-_\Sigma X_i} \and
      \inferrule{\Theta |-_{\Sigma} A_1 \\ \cdots \\ \Theta |-_{\Sigma} A_n}{\Theta |-_{\Sigma} \tyOp_o(A_1, \ldots, A_n)}\;\text{where $n = \arity(o)$}
    \end{mathpar}
    \caption{Simple types}
    \label{fig:simple-type}
  \end{minipage}
  \begin{minipage}[b]{.35\textwidth}
    \centering
    \small
    \judgbox{\Theta |-_{\Sigma} \Gamma}{}
    \begin{mathpar}
      \inferrule{ }{\Theta |-_{\Sigma} \cdot }\and
      \inferrule{\Theta |-_{\Sigma} A \\ \Theta|-_{\Sigma} \Gamma}{\Theta |-_{\Sigma} \Gamma, A}
    \end{mathpar}
    \caption{Contexts}
  \label{fig:simple-context}
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \centering
  \small
  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}}{$\isTerm{t}$ is a raw term for the signature $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{\isTerm{x} : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}}{|-_{\Sigma, \Omega} \isTerm{t \annotate A}}\,\Rule{Anno}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}}
    {|-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op}
  \end{mathpar}
  \begin{flushright}
    for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$
  \end{flushright}%
  \caption{Raw terms}
\end{figure}

\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $\isTerm{t}$ has a type $A$ (without type variables) under $\Gamma$ for the signature $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
    \and
    \inferrule{\Gamma |- \isTerm{t} : A}{\Gamma |- (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
    \and
    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A}{\rho}}\,\Rule{Op}
    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Typing rules}
  \label{fig:extrinsic-typing}
\end{figure}

@@

\begin{figure}
  \centering
  \small
  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$\isTerm{t}$ is a raw term for $(\Sigma, \erase{\Omega})$ in mode $d$ for  $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
    {|-_{\Sigma, \Omega} \tmOpts^\dir{d}}\,\Rule{Op}
  \end{mathpar}
  \caption{Preprocessed raw terms, I}
\end{figure}

\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $\isTerm{t}$ is of type $A$ (without type variables) in mode $\dir{d}$ under $\Gamma$}
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) :^{\syn}  A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ B = A}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
      \cdots \\
    \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A}{\rho}} \,\mathsf{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^\dir{d}$ in $\Omega$}
  \end{mathpar}
Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\chk}}$
  \caption{Bidirectional typing rules}
\end{figure}

### 2023-06-17 tex/sec4-bidirectionalisation.tex
[ tex ] Make some figures more compact
[ tex ] Writing job distribution
[ tex ] Use blackboard font for annotation (in raw terms)
[ tex ] Style raw terms

\subsection{Bidirectionalisation}
\begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
  It is decidable whether a raw term~$t$ is in mode~$d$ for any $t$ and $d$.
\end{proposition}

@@

>>>>

\subsection{Bidirectionalisation}

\begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
  It is decidable whether a raw term~$\isTerm{t}$ is in mode~$d$ for any $\isTerm{t}$ and $d$.
\end{proposition}

@@

\begin{figure}
  \centering
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $\isTerm{t}$ is a raw term possibly with missing annotation or redundant casts in mode $d$ \\
    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $\isTerm{t}$ is a raw term without missing annotation \\
    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $\isTerm{t}$ is a raw term without redundant casts \\
  \end{tabular}
  \egroup
  \LT{Design a notation for \textsf{Pre?} so that a raw term without missing annotation or redundant casts is exactly the intersection of these two cases.}
  \begin{mathpar}
    \inferrule{|-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}\,\SynRule{Anno?}
    \and
    \inferrule{\isTerm{x} : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}
    \and
    \inferrule{|-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn}}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
    {|-_{\Sigma, \Omega} \tmOpts^\dir{d}}
  \end{mathpar}
  \caption{Preprocessed raw terms, II}
  \label{fig:raw-with-missing-annotations}
\end{figure}

\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$ (for some bidirectional type system $(\Sigma, \Omega)$)}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}
              {\isTerm{x} \sqsupseteq \isTerm{x}}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
              {\tmOpts \sqsupseteq \tmOpus}
  \end{mathpar}

  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-order}
\end{figure}
